---
title: "all_st_by_steam"
author: "Rasool Saghaleyni"
date: "2025-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:
```{r, message=FALSE}
library(dplyr)
library(RColorBrewer)
library(scales)
library(reshape2)
library(corrplot)
library(lawstat)
library(MASS)
library(viridis)
library(steam)
```

Load ST data:
```{r}
# 1. Load spatial data (unfiltered)
# se <- Load10X_Spatial(
#   data.dir = "data/Data_brain_example/spatial_data/",
#   filename = "V1_Adult_Mouse_Brain_filtered_feature_bc_matrix.h5",
#   slice = "slice1",
#   assay = "Spatial"
# )
# 
# # 2. Get raw count matrix
# counts <- GetAssayData(se, slot = "counts")
# 
# # 3. Apply gene filters
# genes_to_keep <- rownames(counts)[
#   Matrix::rowSums(counts) >= 100 &     # Total UMI count per gene
#   Matrix::rowSums(counts > 0) >= 5     # Expressed in ≥5 spots
# ]
# 
# # 4. Apply spot (cell) filters
# spots_to_keep <- colnames(counts)[
#   Matrix::colSums(counts) >= 500       # Total UMIs per spot
# ]
# 
# # 5. Subset the Seurat object with filtered genes and spots
# se <- subset(se, features = genes_to_keep, cells = spots_to_keep)
# se
# 
# # alculate mitochondrial and ribosomal percentages
# se$percent.mito <- calculateGenePercentage(se, "^mt-")
# se$percent.ribo <- calculateGenePercentage(se, "^Rpl|^Rps")
# 
# # normalize 
# options(future.globals.maxSize = 4 * 1024^3)
# se <- SCTransform(se, assay = "Spatial", return.only.var.genes = T)
# # add seurat cluster
# seurat_clusters <- read.csv(file = "data/Data_brain_example/spatial_structures/seurat_clusters.csv",
#                            sep = ";",
#                            fileEncoding = "UTF-8",
#                            check.names = FALSE)
# # Clean up barcodes in the cluster annotations to match Seurat object
# seurat_clusters$Barcode <- gsub("_1$", "", seurat_clusters$Barcode)
# 
# se <- addClusterAnnotations(se,
#                            AnnoDataframe = seurat_clusters,
#                            BarcodeName = "Barcode",
#                            AnnoName = "seurat_clusters",
#                            slot_name = "cluster_anno")
# 
# options(future.globals.maxSize = 16 * 1024^3)
# clusterMarkers <- findClusterMarkers(se, nfactors = 40, dims = c(1:22, 24:32, 34:39), M = 10)
# saveRDS(clusterMarkers, file = "data/Data_brain_example/clusterMarkers_nf40_d22_24_32_34_39.rds")
```

Find spatial structures and their top marker genes:
Steps corresponding to `spatial_structure` notebook

```{r}
clusterMarkers <- readRDS("data/Data_brain_example/clusterMarkers_nf40_d22_24_32_34_39.rds")
se <- clusterMarkers$updated_se
all_gene_sets_structure_markers <- clusterMarkers$marker_genes
```


loading OpenTargets gene sets:
Steps corresponding to `OpenTargets_gene_sets` notebook

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
OpenTargets.path <- "data/Data_brain_example/OpenTargets/brain_mouse_10X/MONDO_0004975-associated-diseases_alzheimer_disease_v23.06.tsv"
df <- read.table(paste(OpenTargets.path, sep= "/"),
                     sep= "\t",
                     header=TRUE,
                     quote = "")
column.names <- c("Gene_symbol","Overall_assoc.score","Genetic_assoc.","Somatic_mutation","Drugs","Pathways_system_biology","Text_mining","RNA_expression","Animal_models","Target_name")
df[df == "No data"] <- NA
names(df) <- column.names

mouse_genes = "data/Data_brain_example/OpenTargets_gene_sets/mouse_human_genes_feb_23.tsv"
conversion.table <- read.table(mouse_genes, sep = "\t", header = TRUE)

df$Gene_symbol <- conversion.table[match(df$Gene_symbol,conversion.table$hgnc_symbol),]$mgi_symbol

name = basename(OpenTargets.path)
disease.name <- sub(".*diseases_","", name)
disease.name <- gsub(".tsv", "", disease.name)
df <- df %>% distinct()
#Clean out genes not present in the Seurat object AND removes duplicates
df <- df %>%
  dplyr::select(Gene_symbol, Overall_assoc.score, Genetic_assoc., Drugs, Animal_models) %>%
  distinct() %>%
  filter(!is.na(Gene_symbol) & Gene_symbol %in% rownames(se)) %>%
  group_by(Gene_symbol) %>%
  slice(which.max(Overall_assoc.score)) %>%
  ungroup()
# making Gene sets
GWAS_OpenTarget_Overall_assoc.score_xxx <- arrange(df, desc(Overall_assoc.score))
GWAS_OpenTarget_Overall_assoc.score_xxx <-   GWAS_OpenTarget_Overall_assoc.score_xxx[!is.na(GWAS_OpenTarget_Overall_assoc.score_xxx$Overall_assoc.score),]$Gene_symbol
##genes genetic association score
GWAS_OpenTarget_Genetic_assoc._xxx <- arrange(df, desc(Genetic_assoc.))
GWAS_OpenTarget_Genetic_assoc._xxx <- GWAS_OpenTarget_Genetic_assoc._xxx[!is.na(GWAS_OpenTarget_Genetic_assoc._xxx$Genetic_assoc.),]$Gene_symbol
##genes drug target
GWAS_OpenTarget_Drugs_xxx <- arrange(df, desc(Drugs))
GWAS_OpenTarget_Drugs_xxx <- GWAS_OpenTarget_Drugs_xxx[!is.na(GWAS_OpenTarget_Drugs_xxx$Drugs),]$Gene_symbol

all_gene_sets <- list()
list.names <- character()
all_gene_sets[[1]] <- unique(GWAS_OpenTarget_Overall_assoc.score_xxx)
all_gene_sets[[2]] <- unique(GWAS_OpenTarget_Genetic_assoc._xxx)
all_gene_sets[[3]] <- unique(GWAS_OpenTarget_Drugs_xxx)
list.names <- append(list.names,(c(gsub("xxx", disease.name,"GWAS_OpenTarget_Overall_assoc.score_xxx"),
                            gsub("xxx", disease.name,"GWAS_OpenTarget_Genetic_assoc._xxx"),
                            gsub("xxx", disease.name,"GWAS_OpenTarget_Drugs_xxx"))))
names(all_gene_sets) <- list.names

# Extract Genetic and Drug gene sets
OpenTargets_ALZ_Genetic_ <- all_gene_sets[[grep("Genetic", names(all_gene_sets))]]
OpenTargets_ALZ_Drugs_ <- all_gene_sets[[grep("Drugs", names(all_gene_sets))]]
```


```{r}
window50_rank_list_ALZ_Genetic <- generateSlidingWindows(OpenTargets_ALZ_Genetic_, window_size = 50, step_size = 10, disease_abbr = "ALZ", ot_gene_set_label = "Genetic")

window50_rank_list_ALZ_Drugs <- generateSlidingWindows(OpenTargets_ALZ_Drugs_, window_size = 50, step_size = 10, disease_abbr = "ALZ", ot_gene_set_label = "Drugs")
```


```{r}
OpenTargets_ALZ_Genetic_topGenes_ <- OpenTargets_ALZ_Genetic_[1:50]
```

```{r}
all_gene_sets_OpenTargets <- list(OpenTargets_ALZ_Genetic_=OpenTargets_ALZ_Genetic_,OpenTargets_ALZ_Genetic_topGenes_=OpenTargets_ALZ_Genetic_topGenes_,OpenTargets_ALZ_Drugs_=OpenTargets_ALZ_Drugs_)
```

Merge gene sets:

```{r}
all_gene_sets <- all_gene_sets_OpenTargets
all_gene_sets <- append(all_gene_sets,all_gene_sets_structure_markers)
all_gene_sets <- append(all_gene_sets,window50_rank_list_ALZ_Genetic)
all_gene_sets <- append(all_gene_sets,window50_rank_list_ALZ_Drugs)
```

Calculate the percentage of genes present in the Seurat object:

```{r}
score.mat <- matrix(data = NA, nrow = length(all_gene_sets), ncol = 1, dimnames = list(names(all_gene_sets), "% of genes present"))

for (i in 1:length(all_gene_sets)) {
  genes <- unlist(all_gene_sets[i]) 
  score.mat[i] <- mean(genes %in% rownames(se))
  se <- AddModuleScore(se,list(genes), ctrl = 100, name = list(names(all_gene_sets[i])))
}
```

this chunk takes about 90 minutes for all on my local. Maybe we can further optimize it. So don't try it at home!

```{r}
# Generate permutation matrices (non-parallel, 1000 permutations)

# perm.mat.window50.data <- generatePermutationMatrix(
#   se = se,
#   gnum = 50,
#   permutation_nr = 1000,
#   cluster_anno = "seurat_clusters"
# )
# 
# perm.mat.genetic.data <- generatePermutationMatrix(
#   se = se,
#   gnum = 281,
#   permutation_nr = 1000,
#   cluster_anno = "seurat_clusters"
# )
# 
# perm.mat.drugs.data <- generatePermutationMatrix(
#   se = se,
#   gnum = 132,
#   permutation_nr = 1000,
#   cluster_anno = "seurat_clusters"
# )
```

let's load it for now

```{r}
perm_files <- c(
  "data/Data_brain_example/Permutation_matrices/perm.mat_p10000_g50.RDS",
  "data/Data_brain_example/Permutation_matrices/perm.mat_p10000_g281.RDS",
  "data/Data_brain_example/Permutation_matrices/perm.mat_p10000_g132.RDS"
)

perm_matrices <- loadPermutationMatices(perm_files)
perm.mat.window50.data <- as.data.frame(perm_matrices[[1]])
perm.mat.genetic.data <- as.data.frame(perm_matrices[[2]])
perm.mat.drugs.data <- as.data.frame(perm_matrices[[3]])

colnames(perm.mat.genetic.data) <- gsub("^cluster_", "", colnames(perm.mat.genetic.data))
colnames(perm.mat.window50.data) <- gsub("^cluster_", "", colnames(perm.mat.window50.data))
colnames(perm.mat.drugs.data) <- gsub("^cluster_", "", colnames(perm.mat.drugs.data))
```


Enrichment analysis:


```{r}
p.val.mat.g <- SpatialTraitEnrichmentAnalysis(
  se = se,
  perm.mat.label.data = perm.mat.genetic.data,
  perm.mat.window.data = perm.mat.window50.data,
  window_rank_list_abr_label = window50_rank_list_ALZ_Genetic,
  gene_list = "OpenTargets_ALZ_Genetic_1",
  imputation = "all",
  log_file = "./SpatialTraitEnrichmentAnalysis_log.txt"
)
```

```{r}
# Get all p-values
all_pvals.g <- as.numeric(p.val.mat.g[1, ])

# Adjust p-values (using Bonferroni correction)
adjusted_pvals.g <- p.adjust(all_pvals.g, method = "bonferroni")

# Create a new data frame with both p-values
p.val.mat.g <- rbind(
  p.val.mat.g,
  "p.val.adj" = adjusted_pvals.g
)
```

```{r}
se$seurat_clusters <- as.character(se$seurat_clusters)
clusters <- sort(unique(se$seurat_clusters))

se$p.val.g <- NA
se$p.val.adj.g <- NA

# Match by cluster names
for (cluster in clusters) {
  cluster_name <- paste0("cluster_", cluster)
  if (cluster_name %in% colnames(p.val.mat.g)) {
    se$p.val.g[se$seurat_clusters == cluster] <- as.numeric(p.val.mat.g["p.val", cluster_name])
    se$p.val.adj.g[se$seurat_clusters == cluster] <- as.numeric(p.val.mat.g["p.val.adj", cluster_name])
  } else {
    warning(paste("Cluster", cluster_name, "not found in p.val.mat.g"))
  }
}
```


```{r}
SpatialFeaturePlot(se, 
                  features = "p.val.g",
                  pt.size.factor = 5,
                  image.alpha = 0) +
  scale_color_gradient(low = "white", high = "red")
SpatialFeaturePlot(se, 
                  features = "p.val.adj.g",
                  pt.size.factor = 5,
                  image.alpha = 0) +
  scale_color_gradient(low = "white", high = "red")
```

```{r}
p.val.mat.d <- SpatialTraitEnrichmentAnalysis(
  se = se,
  perm.mat.label.data = perm.mat.genetic.data,
  perm.mat.window.data = perm.mat.window50.data,
  window_rank_list_abr_label = window50_rank_list_ALZ_Genetic,
  gene_list = "OpenTargets_ALZ_Drugs_1",
  imputation = "dynamic",
  log_file = "./SpatialTraitEnrichmentAnalysis_log_st_drugs.txt"
)

# Get all p-values
all_pvals.d <- as.numeric(p.val.mat.d[1, ])

# Adjust p-values (using Bonferroni correction)
adjusted_pvals.d <- p.adjust(all_pvals.d, method = "bonferroni")

# Create a new data frame with both p-values
p.val.mat.d <- rbind(
  p.val.mat.d,
  "p.val.adj" = adjusted_pvals.d
)
```

```{r}
se$seurat_clusters <- as.character(se$seurat_clusters)
clusters <- sort(unique(se$seurat_clusters))

se$p.val.d <- NA
se$p.val.adj.d <- NA

# Match by cluster names
for (cluster in clusters) {
  cluster_name <- paste0("cluster_", cluster)
  if (cluster_name %in% colnames(p.val.mat.d)) {
    se$p.val.d[se$seurat_clusters == cluster] <- as.numeric(p.val.mat.d["p.val", cluster_name])
    se$p.val.adj.d[se$seurat_clusters == cluster] <- as.numeric(p.val.mat.d["p.val.adj", cluster_name])
  } else {
    warning(paste("Cluster", cluster_name, "not found in p.val.mat.d"))
  }
}
```

```{r}
SpatialFeaturePlot(se, 
                  features = "p.val.d",
                  pt.size.factor = 5,
                  image.alpha = 0) +
  scale_color_gradient(low = "white", high = "red")
SpatialFeaturePlot(se, 
                  features = "p.val.adj.d",
                  pt.size.factor = 5,
                  image.alpha = 0) +
  scale_color_gradient(low = "white", high = "red")
```


```{r}
window_results <- WindowRankEnrichmentAnalysis(
    se = se,
    perm.mat.window.data = perm.mat.window50.data,
    window_rank_list = window50_rank_list_ALZ_Drugs,
    ot_gene_set_label = "Drugs",
    disease_abbr = "ALZ"
)

# adjust p-values
window_results$p_value_adj <- p.adjust(window_results$p_value, method = "bonferroni")
# other methods:
# window_results$p_value_adj <- p.adjust(window_results$p_value, method = "fdr")
# window_results$p_value_adj <- p.adjust(window_results$p_value, method = "holm")
```


```{r}
rankPlotForClusters(
    window_results = window_results,
    perm.mat.window = perm.mat.window50.data,
    ylim = c(-0.3, 0.3)
)
```


```{r}
featurePlotForRanks(se = se,
                   window_rank_list = window50_rank_list_ALZ_Genetic,
                   ot_gene_set_label = "Genetic",
                   disease_abbr = "ALZ",
                   cluster_number = NULL, 
                   ranks_per_plot = 1,
                   spot_alpha = 1, 
                   point_size = 5,
                   #seurat_type = "seurat",
                   all_clusters = TRUE)

```


```{r}
plotScoreDist(
  se = se,
  perm.mat = perm.mat.genetic.data,
  cluster_anno = "seurat_clusters",
  enrichment_score_col = "OpenTargets_ALZ_Genetic_1"
)
```

```{r}
colnames(perm.mat.genetic.data) <- gsub("^cluster_", "", colnames(perm.mat.genetic.data))
colnames(perm.mat.window50.data) <- gsub("^cluster_", "", colnames(perm.mat.window50.data))
plotScoreDist(
  se = se,
  perm.mat = perm.mat.drugs.data,
  cluster_anno = "seurat_clusters",
  enrichment_score_col = "OpenTargets_ALZ_Drugs_1"
)
```
