---
title: "all_sc_by_steam"
author: "Rasool Saghaleyni"
date: "2025-04-11"
output: html_document
---

Load libraries
```{r}
library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyverse)
library(steam)
library(ggplot2)
library(viridis)
```


Load SC se object:
```{r}
# se <- readRDS("data/Seurat_objects/se_allen_brain_atlas_random_250_subclass_label.rds")
# se
# table(se$subclass_label)
```


Collect QC metrics
```{r}
# se[["percent.mt"]] <- PercentageFeatureSet(se, pattern = "^mt-")
# se[["percent.rb"]] <- PercentageFeatureSet(se, pattern = "^Rpl|^Rps")
```

Filtering cells
```{r}
# se
# se <- subset(se, subset = nFeature_RNA > 200 & nFeature_RNA < 8000 &
#                     nCount_RNA > 500 & nCount_RNA < 35000 &
#                     percent.mt <= 10 & percent.rb <= 15)
# se
```

Filtering genes
```{r}
# se
# # Define a threshold for the total UMI count
# min_umi <- 100
# 
# # Calculate the total UMIs per gene
# gene_totals <- Matrix::rowSums(GetAssayData(se, slot = "counts"))
# 
# # Subset genes with total UMIs above the threshold
# se <- subset(se, features = rownames(se)[gene_totals >= min_umi])
# se
```

Save cluster annotations
```{r, eval=FALSE}
# seurat_clusters <- as.data.frame(se$subclass_label)
# seurat_clusters$Barcode <- colnames(se)
# seurat_clusters$seurat_clusters <- seurat_clusters$`se$subclass_label`
# seurat_clusters$`se$subclass_label` <- NULL
# write.table(seurat_clusters,file="data/Seurat_objects/seurat_clusters.csv", col.names = T, row.names = F,quote = FALSE,sep = ";")
```

Normalize data: se <- SCTransform(se,return.only.var.genes = T) and Save filtered and normalized se object

```{r, message=FALSE, include=FALSE}
# options(future.globals.maxSize = 5 * 1024^3)
# se <- SCTransform(se,return.only.var.genes = T)
# saveRDS(se,file = "data/Seurat_objects/se_250_cells_filtered_normalized.RDS")
# se <- readRDS("data/Seurat_objects/se_250_cells_filtered_normalized.RDS")
```

Generate metacell clusters

```{r}
# clustered_metacells <- generateMetacellClusters(se, cluster_anno = "subclass_label", verbose = FALSE)
```

Aggregate and save metacell clusters

```{r}
# meta_cell_seurat_all_sum_with_metadata <- aggregateMetacells(clustered_metacells)
# saveRDS(meta_cell_seurat_all_sum_with_metadata,file("data/Seurat_objects/MetaCell.rds"))
```

Read saved aggregated MetaCell Seurat Object

```{r,eval=T}
# meta_cell_seurat_all_sum_with_metadata <- readRDS("data/Seurat_objects/MetaCell.rds")

```

Save cluster annotations
```{r, eval=T}
# seurat_clusters <- as.data.frame(meta_cell_seurat_all_sum_with_metadata$subclass_label)
# seurat_clusters$Barcode <- colnames(meta_cell_seurat_all_sum_with_metadata)
# seurat_clusters$seurat_clusters <- seurat_clusters$`meta_cell_seurat_all_sum_with_metadata$subclass_label`
# seurat_clusters$`meta_cell_seurat_all_sum_with_metadata$subclass_label` <- NULL
# write.table(seurat_clusters,file="data/Seurat_objects/seurat_clusters.csv", col.names = T, row.names = F,quote = FALSE,sep = ";")

seurat_clusters <- read.table("data/Seurat_objects/seurat_clusters.csv",
                               header = TRUE,
                               sep = ";",
                               stringsAsFactors = FALSE)

```


```{r, fig.width=35, fig.height=15}
se <- readRDS("data/Seurat_objects/MetaCell_normalized.RDS")
se <- SetIdent(se, value = "supercluster_term")
```

Load OpenTargets gene sets and make a list of gene sets

```{r class.source = 'fold-show'}

OpenTargets_SCZ_Genetic_ <- as.character(t(read.table(file = "data/OpenTargets_gene_sets/Human_OpenTargets_SCZ_Genetic.csv")))
length(OpenTargets_SCZ_Genetic_)

OpenTargets_SCZ_Drugs_ <- as.character(t(read.table(file = "data/OpenTargets_gene_sets/Human_OpenTargets_SCZ_Drugs.csv")))
length(OpenTargets_SCZ_Drugs_)
all_gene_sets_OpenTargets <- list(OpenTargets_SCZ_Genetic_=OpenTargets_SCZ_Genetic_,OpenTargets_SCZ_Drugs_=OpenTargets_SCZ_Drugs_)

```

Load window rank lists
```{r}
window50_rank_list_SCZ_Genetic <- readRDS(file = "data/OpenTargets_windowLists/Human_window50_rank_list_SCZ_Genetic.RDS")
window50_rank_list_SCZ_Drugs <- readRDS(file = "data/OpenTargets_windowLists/Human_window50_rank_list_SCZ_Drugs.RDS")
```

Merge gene sets
```{r}
all_gene_sets <- all_gene_sets_OpenTargets
all_gene_sets <- append(all_gene_sets,window50_rank_list_SCZ_Genetic)
all_gene_sets <- append(all_gene_sets,window50_rank_list_SCZ_Drugs)
```

# Score data based on gene sets

AddModuleScore loop se:
```{r, results='hide'}

score.mat <- matrix(data = NA, nrow = length(all_gene_sets), ncol = 1, dimnames = list(names(all_gene_sets), "% of genes present"))

for (i in 1:length(all_gene_sets)) {
  genes <- unlist(all_gene_sets[i]) # character vector of genes in a gene set
  score.mat[i] <- mean(genes %in% rownames(se)) # calculate and collect % of genes present in se from the gene set
  se <- AddModuleScore(se,list(genes), ctrl = 100, name = list(names(all_gene_sets[i])))
}

 score.mat*100
```


# NULL set permutation

```{r, eval=FALSE}
#generate permutation matrices (non-parallel, 1000 permutations)

# perm.mat_p1000_g50 <- generatePermutationMatrix(
#   se = se,
#   gnum = 50,
#   permutation_nr = 1000,
#   cluster_anno = "supercluster_term"
# )
# 
# saveRDS(perm.mat_p1000_g50, file = "Permutation_matrices/perm.mat_p1000_g50.RDS")
# write.matrix(perm.mat_p1000_g50, file = "Permutation_matrices/perm.mat_p1000_g50.csv", sep = ";")
# 
# perm.mat_p1000_g218 <- generatePermutationMatrix(
#   se = se,
#   gnum = 218,
#   permutation_nr = 1000,
#   cluster_anno = "supercluster_term"
# )
# 
# saveRDS(perm.mat_p1000_g218, file = "Permutation_matrices/perm.mat_p1000_g218.RDS")
# write.matrix(perm.mat_p1000_g218, file = "Permutation_matrices/perm.mat_p1000_g218.csv", sep = ";")
# 
# perm.mat_p1000_g1315 <- generatePermutationMatrix(
#   se = se,
#   gnum = 1315,
#   permutation_nr = 1000,
#   cluster_anno = "supercluster_term"
# )
# 
# saveRDS(perm.mat_p1000_g1315, file = "Data_brain_example/Permutation_matrices/perm.mat_p1000_g1315.RDS")
# write.matrix(perm.mat_p1000_g1315, file = "Data_brain_example/Permutation_matrices/perm.mat_p1000_g1315.csv", sep = ";")

```

Load perm.mat

```{r}
perm_files <- c(
  "data/Permutation_matrices/perm.mat_p10000_g50.RDS",
  "data/Permutation_matrices/perm.mat_p10000_g1315.RDS",
  "data/Permutation_matrices/perm.mat_p10000_g218.RDS"
)

perm_matrices <- loadPermutationMatices(perm_files)
perm.mat.window50.data <- as.data.frame(perm_matrices[[1]])
perm.mat.genetic.data <- as.data.frame(perm_matrices[[2]])
perm.mat.drugs.data <- as.data.frame(perm_matrices[[3]])
```

Genetics:

```{r}
p.val.mat.g <- SpatialTraitEnrichmentAnalysis(
  se = se,
  perm.mat.label.data = perm.mat.genetic.data,
  perm.mat.window50.data = perm.mat.window50.data,
  window_rank_list_abr_label = window50_rank_list_SCZ_Genetic,
  gene_list = "OpenTargets_SCZ_Genetic_1",
  cluster_anno = "supercluster_term",
  imputation = "dynamic",
  log_file = "./SpatialTraitEnrichmentAnalysis_all_sc_log.txt"
)

# Get all p-values
all_pvals.g <- as.numeric(p.val.mat.g[1, ])

# Adjust p-values (using Bonferroni correction)
adjusted_pvals.g <- p.adjust(all_pvals.g, method = "bonferroni")

# Create a new data frame with both p-values
p.val.mat.g <- rbind(
  p.val.mat.g,
  "p.val.adj" = adjusted_pvals.g
)
```

```{r}
se$supercluster_term <- as.character(se$supercluster_term)
clusters <- sort(unique(se$supercluster_term))

se$p.val.g <- NA
se$p.val.adj.g <- NA

# Match by cluster names
for (cluster in clusters) {
  cluster_name <- paste0("cluster_", cluster)
  if (cluster_name %in% colnames(p.val.mat.g)) {
    se$p.val.g[se$supercluster_term == cluster] <- as.numeric(p.val.mat.g["p.val", cluster_name])
    se$p.val.adj.g[se$supercluster_term == cluster] <- as.numeric(p.val.mat.g["p.val.adj", cluster_name])
  } else {
    warning(paste("Cluster", cluster_name, "not found in p.val.mat.g"))
  }
}
```

```{r}
# Ensure supercluster_term is character
# se$supercluster_term <- as.character(se$supercluster_term)
# 
# se@meta.data$p.val.g <- setNames(as.numeric(p.val.mat.g[1, ]), gsub("^cluster_", "", colnames(p.val.mat.g)))[se$supercluster_term]
# se@meta.data$p.val.adj.g <- setNames(as.numeric(p.val.mat.g[2, ]), gsub("^cluster_", "", colnames(p.val.mat.g)))[se$supercluster_term]
```

Drugs:

```{r}
p.val.mat.d <- SpatialTraitEnrichmentAnalysis(
  se = se,
  perm.mat.label.data = perm.mat.genetic.data,
  perm.mat.window50.data = perm.mat.window50.data,
  window_rank_list_abr_label = window50_rank_list_SCZ_Genetic,
  gene_list = "OpenTargets_SCZ_Drugs_1",
  cluster_anno = "supercluster_term",
  imputation = "dynamic"
)

# Get all p-values
all_pvals.d <- as.numeric(p.val.mat.d[1, ])

# Adjust p-values (using Bonferroni correction)
adjusted_pvals.d <- p.adjust(all_pvals.d, method = "bonferroni")

# Create a new data frame with both p-values
p.val.mat.d <- rbind(
  p.val.mat.d,
  "p.val.adj" = adjusted_pvals.d
)
```

```{r}
se$supercluster_term <- as.character(se$supercluster_term)
clusters <- sort(unique(se$supercluster_term))

se$p.val.d <- NA
se$p.val.adj.d <- NA

# Match by cluster names
for (cluster in clusters) {
  cluster_name <- paste0("cluster_", cluster)
  if (cluster_name %in% colnames(p.val.mat.d)) {
    se$p.val.d[se$supercluster_term == cluster] <- as.numeric(p.val.mat.d["p.val", cluster_name])
    se$p.val.adj.d[se$supercluster_term == cluster] <- as.numeric(p.val.mat.d["p.val.adj", cluster_name])
  } else {
    warning(paste("Cluster", cluster_name, "not found in p.val.mat.d"))
  }
}
```

```{r}
# se@meta.data$p.val.d <- setNames(as.numeric(p.val.mat.d[1, ]), gsub("^cluster_", "", colnames(p.val.mat.d)))[se$supercluster_term]
# se@meta.data$p.val.adj.d <- setNames(as.numeric(p.val.mat.d[2, ]), gsub("^cluster_", "", colnames(p.val.mat.d)))[se$supercluster_term]
```

### Violin plots 1-5
```{r, fig.width=35, fig.height=15}
se <- SetIdent(se, value = "supercluster_term")

VlnPlot(se, features = c("SCZ_Genetic_Rank1_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Genetic_Rank2_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Genetic_Rank3_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Genetic_Rank4_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Genetic_Rank5_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))

```

```{r}
window_results.g <- WindowRankEnrichmentAnalysis(
    se = se,
    perm.mat.window50.data = perm.mat.window50.data,
    window_rank_list = window50_rank_list_SCZ_Genetic,
    ot_gene_set_label = "Genetic",
    disease_abbr = "SCZ",
    cluster_anno = "supercluster_term"
)
```

```{r}
rankPlotForClusters(
    window_results = window_results.g,
    perm.mat.window50 = perm.mat.window50.data,
    ylim = c(-0.3, 0.3),
    clusters_to_plot = NULL
)
```


```{r}
plotScoreDist(
  se = se,
  perm.mat = perm.mat.genetic.data,
  perm.mat.window50 = perm.mat.window50.data,
  window_rank_list = window50_rank_list_SCZ_Genetic,
  cluster_anno = "supercluster_term",
  enrichment_score_col = "OpenTargets_SCZ_Genetic_1"
)

```

### Violin plots 1-5
```{r, fig.width=35, fig.height=15}
se <- SetIdent(se, value = "supercluster_term")

VlnPlot(se, features = c("SCZ_Drugs_Rank1_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Drugs_Rank2_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Drugs_Rank3_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Drugs_Rank4_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
VlnPlot(se, features = c("SCZ_Drugs_Rank5_1"), ncol = 1, group.by = "supercluster_term") + geom_boxplot(width=0.4) + theme(plot.title = element_text(size = 12))
```

```{r}
window_results_drugs <- WindowRankEnrichmentAnalysis(
    se = se,
    perm.mat.window50.data = perm.mat.window50.data,
    window_rank_list = window50_rank_list_SCZ_Drugs,
    ot_gene_set_label = "Drugs",
    disease_abbr = "SCZ",
    cluster_anno = "supercluster_term"
)
```


```{r}
rankPlotForClusters(
    window_results = window_results_drugs,
    perm.mat.window50 = perm.mat.window50.data,
    ylim = c(-0.3, 0.3),
    clusters_to_plot = NULL
)
```

```{r}
plotScoreDist(
  se = se,
  perm.mat = perm.mat.drugs.data,
  perm.mat.window50 = perm.mat.window50.data,
  window_rank_list = window50_rank_list_SCZ_Drugs,
  cluster_anno = "supercluster_term",
  enrichment_score_col = "OpenTargets_SCZ_Drugs_1"
)

```

# Genetic p-value plot {.tabset}

```{r}
# Create empty vectors
se$neg.log10.p.val.g <- NA
se$neg.log10.p.val.g.adj <- NA

# Loop over clusters and assign values from p.val.mat.g
for (i in seq_along(clusters)) {
  se$neg.log10.p.val.g[se$supercluster_term == clusters[i]] <- p.val.mat.g[1, i]
  se$neg.log10.p.val.g.adj[se$supercluster_term == clusters[i]] <- p.val.mat.g[2, i]
}
```

```{r}
genetic_meta.data <- se@meta.data %>%
  tibble::rownames_to_column("CellID") %>%
  dplyr::select(CellID, supercluster_term, neg.log10.p.val.g, neg.log10.p.val.g.adj)

# Convert to numeric (in case of factor/string)
genetic_meta.data$neg.log10.p.val.g <- as.numeric(genetic_meta.data$neg.log10.p.val.g)
genetic_meta.data$neg.log10.p.val.g.adj <- as.numeric(genetic_meta.data$neg.log10.p.val.g.adj)
```

```{r}
genetic_meta.data <- genetic_meta.data %>%
  mutate(
    significance.level = case_when(
      10^(-neg.log10.p.val.g.adj) < 0.001 ~ "***",
      10^(-neg.log10.p.val.g.adj) < 0.01  ~ "**",
      10^(-neg.log10.p.val.g.adj) < 0.05  ~ "*",
      TRUE ~ ""
    ),
    significant = ifelse(10^(-neg.log10.p.val.g.adj) < 0.05, "TRUE", "FALSE"),
    disease = "SCZ"
  )

```

```{r, fig.width=20, fig.height=10}
p <- genetic_meta.data %>%
  ggplot(aes(x = supercluster_term, y = disease, fill = significant))+
  geom_tile(colour = "grey20")+
  scale_fill_manual(values = c("grey75","red"), breaks = c("FALSE", "TRUE"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.5, "cm"))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 22))+
  coord_equal()+
  labs(x = "cell type",
       fill = "nominal p-value\n significant")


p
```

```{r, fig.width=20, fig.height=10}

#genetic_meta.data$p.val.g <- as.numeric(genetic_meta.data$p.val.g)

p <- genetic_meta.data %>%
  ggplot(aes(x = supercluster_term, y = disease)) +
  geom_tile(aes(color = significance.level)) +
  geom_tile(colour = "grey20", aes(fill = neg.log10.p.val.g)) +
  geom_text(data = genetic_meta.data %>% filter(significant == TRUE),
            aes(label = significance.level),
            vjust = 0.75,
            size = 5,
            fontface = "bold") +
  theme(
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(0.5, "cm"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold", size = 10),
    axis.text.y = element_text(face = "bold", size = 10)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 22)) +
  coord_equal() +
  labs(x = "cell type",
       fill = "nominal p-value\n -log10(p-value)") +
  scale_fill_viridis_c(begin = 0.2, option = "viridis") +
  scale_color_manual(values = c("*" = "white", "**" = "white", "***" = "white"),  
                     labels = c("***" = "***  p < 0.001", "**" = "**  p < 0.01", "*" = "*  p < 0.05"),
                     guide = guide_legend(override.aes = list(shape = 21,  
                                                              fill = "white",
                                                              size = 4)))

p

```


# Drugs p-value plot {.tabset}

```{r}
# Initialize metadata columns
se$neg.log10.p.val.d <- NA
se$neg.log10.p.val.d.adj <- NA

# Get ordered cluster labels
clusters <- names(table(unique(se$supercluster_term)))

# Assign values from p.val.mat.d using a loop
for (i in seq_along(clusters)) {
  se$neg.log10.p.val.d[se$supercluster_term == clusters[i]] <- p.val.mat.d[1, i]
  se$neg.log10.p.val.d.adj[se$supercluster_term == clusters[i]] <- p.val.mat.d[2, i]
}
```

```{r}
drugs_meta.data <- se@meta.data %>%
  tibble::rownames_to_column("CellID") %>%
  dplyr::select(CellID, supercluster_term, neg.log10.p.val.d, neg.log10.p.val.d.adj) %>%
  mutate(
    neg.log10.p.val.d = as.numeric(neg.log10.p.val.d),
    neg.log10.p.val.d.adj = as.numeric(neg.log10.p.val.d.adj)
  )

```

```{r}
drugs_meta.data <- drugs_meta.data %>%
  mutate(
    significance.level = case_when(
      10^(-neg.log10.p.val.d.adj) < 0.001 ~ "***",
      10^(-neg.log10.p.val.d.adj) < 0.01  ~ "**",
      10^(-neg.log10.p.val.d.adj) < 0.05  ~ "*",
      TRUE ~ ""
    ),
    significant = ifelse(10^(-neg.log10.p.val.d.adj) < 0.05, "TRUE", "FALSE"),
    disease = "SCZ"
  )

```

```{r, fig.width=20, fig.height=10}
p <- drugs_meta.data %>%
  ggplot(aes(x = supercluster_term, y = disease, fill = significant))+
  geom_tile(colour = "grey20")+
  scale_fill_manual(values = c("grey75","red"), breaks = c("FALSE", "TRUE"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.5, "cm"))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 22))+
  coord_equal()+
  labs(x = "cell type",
       fill = "nominal p-value\n significant")


p
```

```{r, fig.width=20, fig.height=10}
#drugs_meta.data$p.val.d <- as.numeric(drugs_meta.data$p.val.d)

p <- drugs_meta.data %>%
  ggplot(aes(x = supercluster_term, y = disease)) +
  geom_tile(aes(color = significance.level)) +
  geom_tile(colour = "grey20", aes(fill = neg.log10.p.val.d)) +
  
  geom_text(data = drugs_meta.data %>% filter(significant == TRUE),
            aes(label = significance.level),
            vjust = 0.75,
            size = 5,
            fontface = "bold") +

  theme(
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(0.5, "cm"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, face = "bold", size = 10),
    axis.text.y = element_text(face = "bold", size = 10)
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 22)) +
  coord_equal() +
  labs(x = "cell type",
       fill = "nominal p-value\n -log10(p-value)") +
  
  scale_fill_viridis_c(begin = 0.2, option = "viridis") +
  
  scale_color_manual(values = c("*" = "white", "**" = "white", "***" = "white"),  
                     labels = c("***" = "***  p < 0.001", "**" = "**  p < 0.01", "*" = "*  p < 0.05"),
                     guide = guide_legend(override.aes = list(
                       shape = 21,  
                       fill = "white",
                       size = 4)))
p

```

```{r}
# DefaultAssay(se) <- "RNA"
# se <- se %>%
#   NormalizeData() %>%
#   ScaleData()
# 
# se <- SetIdent(se, value = "supercluster_term")
# N.markers <- FindAllMarkers(se, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25,slot = "data")
# write.csv(N.markers, file = "data/DEG_output/DE_genes.csv", row.names = FALSE, quote = FALSE)
```

```{r}
DEG_genes <- read.csv("data/DEG_output/DE_genes.csv", stringsAsFactors = FALSE)
DEG_genes <- as.data.frame(DEG_genes)
N.markers <- DEG_genes
```

```{r}
topN <- N.markers %>% 
            group_by(cluster) %>% 
            arrange(cluster,p_val_adj) %>%
            slice_head(n=10)
```

```{r}
DoHeatmap(se, features = topN$gene)

```

```{r}
OpenTargets_SCZ_Genetic <- OpenTargets_SCZ_Genetic_[1:290]

```

```{r}
# View results
fisher_results <- runFisherTestDEGs(se, DEG_genes = N.markers, OT_genes = OpenTargets_SCZ_Genetic)
print(fisher_results)
```


# Plot results
```{r, fig.height=5, fig.width=25}
# Ensure clusters are ordered by significance
fisher_results$cluster <- factor(fisher_results$cluster, 
                                 levels = fisher_results$cluster[order(fisher_results$adj_p_value)])

# Create the updated plot with viridis scale
p1 <- ggplot(fisher_results, aes(x = cluster, color = -log10(p_value), y = -log10(adj_p_value), size = odds_ratio)) +
  geom_point() +  
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +  
  scale_color_viridis(option = "D", direction = 1, name = "-log10(p-value)") +  # Use Viridis color scale
  scale_size_continuous(range = c(2, 8), name = "Odds ratio") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size = 14),axis.text.y = element_text(size = 14),axis.title.x = element_text(size = 16, face = "bold"),axis.title.y = element_text(size = 16, face = "bold"),plot.title = element_text(size = 18, face = "bold"),legend.title = element_text(size = 14),legend.text = element_text(size = 12)) +  
  labs(title = "Enrichment of top genetically associated SCZ genes",
       x = "Cell types",
       y = "-log10(adj p-value)") +
  guides(color = guide_colorbar(order = 1), size = guide_legend(order = 2))

p1
```
















